---
title: R package *nplr* \n *n-parameter logistic regressions*
author: "Frederic Commo, Briant M. Bot"
output: rmarkdown::html_vignette
bibliography: references.bib  
csl: ieee.csl
vignette: >
  %\VignetteIndexEntry{R package *nplr* \n *n-parameter logistic regressions*}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
# Introduction
  
## Overview
In in-vitro experiments, the aim of drug response analyses is usually to estimate the drug concentration required to reach a given cell line growth inhibition rate—typically the 50% inhibitory concentration ($IC_{50}$), which inhibits 50% of the proliferation, compared with an untreated control. This estimation can be achieved by modeling the inhibition rate observed under a range of drug concentrations. Once the model is fitted, the x values (drug concentrations) can be estimated from the y values (inhibition rates) by simply inverting the function.

The most commonly used model for drug response analysis is the Richards' equation [@Richards], also referred to as a 5-parameter logistic regression [@Giraldo]:

$$y = B + \dfrac{T - B}{\left [1 + 10^{b(x_{mid} - x)} \right ]^s}$$

where $B$ and $T$ are the bottom and top asymptotes, and $b$, $x_{mid}$, and $s$ are the Hill slope, the x-coordinate at the inflection point, and an asymmetric coefficient, respectively.

The *nplr* package is based on the full 5-parameter model, where all of the parameters are optimized, simultaneously, using a Newton-Raphson method (`nlm`, R package `stats`). The objective function to minimize is a weighted sum of squared errors:

$$ sse(Y) = \Sigma_i{w_i.(\hat{y}_i - y_i)^2}, i=1,...,n $$

The weights, $w_i$, used in the objective function can be computed using 3 possible methods, as follows:

- **Residuals weights:** $w_i = \left (\frac{1}{res_i}\right )^p, i=1,...,n$ values
- **Standard weights:** $w_{ir} = \frac{1}{Var(y_r)}, r=1,...,r$ replicated conditions
- **General weights:** $w_{i} = \left (\frac{1}{\hat{y}_i}\right )^p, i=1,...,n$ values

where $p$ is a tuning parameter. The *nplr* package provides several options for computing flexible weighted n-parameter logistic regression: `npars="all"` can be explicitly specified, from 2 to 5, or set to `all`. In that case, all the possible models are evaluated, and the optimal one is returned with respect to the minimal error returned by `nlm`.

The final model performance is estimated by a weighted and non-weighted standard error, as well as the weighted and non-weighted goodness-of-fit:

- **Standard error:** $ \frac{1}{(n-2)}.\Sigma_i{(\hat{y}_i - y_i)^2}, i=1,...,n $
- **Weighted standard error:** $ \frac{1}{(n-2)}.\Sigma_i{w_i.(\hat{y}_i - y_i)^2}, i=1,...,n $
- **Goodness-of-fit:** $ 1 - \frac{SS_{res}}{SS_{tot}} $
- **Weighted goodness-of-fit:** $ 1 - \frac{\Sigma_i{w_i(\hat{y}_i - y_i)^2}}{\Sigma_i{(y_i - \overline{y})^2}} $

## Functions in *nplr*

The main function is simply `nplr()`. It requires 2 main arguments: a vector of `x` and a vector of `y`. Several other arguments have default values.

The `npars` argument allows a user to run specific n-parameter models, n from 2 to 5, while the default value, `npars="all"`, asks the function to test which model fits the best the data, according to a weighted Goodness-of-Fit estimator. In some situations, the x values may need to be log-transformed, e.g. x is provided as original drug concentrations. In such cases, setting `useLog=TRUE` in `nplr()` will apply a $Log_{10}$ transformation on the x values.

The `nplr()` function has been optimized for fitting curves on y-values passed as proportions of control, between 0 to 1. If data are supplied as original response values, e.g. optical density measurements, the `convertToProp()` function may be helpful. In drug-response curve fitting, a good practice consists in adjusting the signals on a $T_0$ and a $control$ (Ctrl) values. Providing these values, the proportion values, $y_p$, are computed as:

$$y_p = \frac{y - T_0}{Ctrl - T_0}$$

where $y$, $T_0$, and $Ctrl$ are the observed values, the 'time zero', and the 'untreated control', respectively.

Note that if neither $T_0$ nor $Ctrl$ are provided, `convertToProp()` will compute the proportions with respect to the $min$ and $max$ of $y$. In that case, the user should be aware that $y = 0.5$ may not correspond to a $IC_{50}$, but rather to a $EC_{50}$ (the half-effect between the maximum and the minimum of the observed effects).

In a drug-response (or progression) curve fitting context, typical needs are to invert the function in order to estimate the x value, e.g. the $IC_50$, given a $y$ value, e.g. the 0.5 survival rate. To do so, the implemented `getEstimates()` method takes 2 arguments: the model (an instance of the class nplr), and one (or a vector of) target(s). `getEstimates()` returns the corresponding x values and their estimated confidence intervals, as specified by `conf.level`.

# Examples

The examples below use some samples of the NCI-60 Growth Inhibition Data. The full data can be downloaded at [NCI](https://wiki.nci.nih.gov/display/NCIDTPdata/NCI-60+Growth+Inhibition+Data). For the purpose of the demonstration, the supplied drug concentrations have been re-exponentiated.

## Example 1

### Fitting a model

```{r nplr, message=FALSE, warning=FALSE}
require(nplr)
```

The first example fits a simple drug-response curve: the PC-3 cell line treated with Thioguanine, 19 points without replicates.

```{r test1}
path <- system.file("extdata", "pc3.txt", package="nplr")
pc3 <- read.delim(path)
np1 <- nplr(x=pc3$CONC, y=pc3$GIPROP)
```

Calling the object returns the fitting summary for the model.

```{r out1}
np1
```

### Visualizing the model

A specific `plot()` function has been implemented in order to visualize the results.

```{r plot1, include=FALSE}
plot(np1, cex.main = 1.2, main="PC-3 cell line. Response to Thioguanine")
```

```{r fig1, ref.label='plot1'}
```

This function has several predefined graphical parameters, and some of them can be overwritten. However, a convenient way to draw simplest or customized plots is shown in the example below:

```{r custom, include=FALSE}
op <- par(no.readonly=TRUE)
par(mfrow=c(1,2))
plot(np1, pcol="grey40", lcol="skyblue1", showEstim=.5, showInfl=TRUE,
     main="Default 'nplr' plot", cex.main=1.5)
x1 <- getX(np1); y1 <- getY(np1)
x2 <- getXcurve(np1); y2 <- getYcurve(np1)
plot(x1, y1, pch=15, cex=2, col="tan1", xlab=expression(Log[10](conc)),
     ylab="Prop", main="Custom plot", cex.main=1.5)
lines(x2, y2, lwd=5, col="seagreen4")
par(op)
```

```{r plotCustom, ref.label='custom', fig.width=12, fig.height=5.5, echo=3:7}
```

### Accessing the performances

Once the model is built, several accessor functions allow to get access to the performances of the model, and its parameters.

```{r getGoodness1}
getGoodness(np1)
```

```{r getStdErr1}
getStdErr(np1)
```

```{r getPar1}
getPar(np1)
```

Here, the 5-parameter model has been chosen as it showed better performances, according to the goodness-of-fit (`npar=5`). The optimal values for the parameters are reported in `params`.

### Estimating the drug concentrations

The purpose of such fitting is to estimate the response to the drug. To do so, `nplr` provides 2 estimates: the area under the curve (AUC), and the drug concentration for a given response to reach.

The `getAUC()` function returns the area under the curve (AUC) estimated by

 integration of the logistic function. This can be useful for further analysis of the pharmacological profile of the drug under study.

```{r getAUC1}
getAUC(np1)
```

The function `getEstimates()` provides estimates of the concentrations for a given response. In this case, we can estimate the concentration at 50% inhibition rate.

```{r getEstimates1}
getEstimates(np1, 0.5)
```

## Example 2

### Fitting a model

For the second example, we use the same NCI-60 Growth Inhibition Data, with a 3-parameter model.

```{r test2}
np2 <- nplr(x=pc3$CONC, y=pc3$GIPROP, npars=3)
```

```{r np2}
np2
```

```{r plot2, include=FALSE}
plot(np2, cex.main = 1.2, main="PC-3 cell line. Response to Thioguanine (3-parameters model)")
```

```{r fig2, ref.label='plot2'}
```

## Example 3

### Fitting a model

For the third example, we use the same NCI-60 Growth Inhibition Data, but with the default `npars="all"`. This will fit all possible models (2 to 5 parameters), and returns the best fit based on the minimal goodness-of-fit score.

```{r test3}
np3 <- nplr(x=pc3$CONC, y=pc3$GIPROP, npars="all")
```

```{r np3}
np3
```

```{r plot3, include=FALSE}
plot(np3, cex.main = 1.2, main="PC-3 cell line. Response to Thioguanine (Best model)")
```

```{r fig3, ref.label='plot3'}
```

## Conclusion

The *nplr* package provides a comprehensive and flexible framework for performing n-parameter logistic regression to analyze drug response data. The package supports flexible weighting of data, automated model selection, and estimation of key parameters such as IC50. Through its simple and intuitive interface, users can easily fit, visualize, and assess the performance of drug-response models.
